Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-11-24 demystify.tex
updates

\usepackage[T1]{fontenc}
%\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
%\usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=blue,
    urlcolor=blue,
}
\usepackage{booktabs}

@@

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    %\hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
  }
\makeatother
%\author{ThanhVu H. Nguyen}
%\title{TITLE OF BOOK}
%\date{}

@@

\chapter{Preface}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students} (and even domestic ones) lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

%though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines. 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and also likely universities in other countries).

@@

\chapter{Summary}\label{sec:summary}
Here we summarize the main points of the document. This allows you to quickly skim through the document and decide which parts you want to read in detail.
\begin{enumerate}
\item Should you apply to a CS PhD in the US?
\begin{itemize}
    \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
\end{itemize}
\item How is your application evaluated?
\begin{itemize}
    \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
    \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
\end{itemize}
\item Application Materials
\begin{itemize}
    \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
    \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
    \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
    \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
    \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
    \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
\end{itemize}
\item What to do after getting admitted?
\begin{itemize}
    \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
\end{itemize}
\item Funding
\begin{itemize}
    \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
\end{itemize}
\item Choosing School and Professors
\begin{itemize}
    \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
    \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
\end{itemize}
\item Miscs and FAQS
\begin{itemize}
    \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
    \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
    \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
    \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
    \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
\end{itemize}
\end{enumerate}

@@

%\didi{Is there more information on typical strengths and common weaknesses of applications. This is especially useful to sophomore and junior students as they still have time to work on those strengths.}
% tvn: the main thing is research experiences

@@

%\begin{commentbox}[Hung]
%    The application fee varies between schools, and is about \$70--\$100. I found this number pretty ridiculous. In most cases, there is no proper justification for that amount. Yes, the application fee could act as a filter, but a nominal amount would do the job. While a blanket waiver might not be feasible, universities should extend the eligibility of fee waiver applications to more countries.
%\end{commentbox}

@@

It's important to aim for the top venues in your field. You can
find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \S\ref{sec:ranking}). Local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty.

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general confs/journals, even those under submissions or rejected, would suffice.  But be sure to upload your papers with your application and talk about them in your statement (see \S\ref{sec:research-statement}).

@@

CS PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, mentors) that know you well enough to talk about \emph{your specific research experience and capabilities}. This person can be a postdoc or even graduate student who has mentored you.

@@

For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the US. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters can be positive as they have helped many students getting into top PhD programs in the US.

@@

@@

    \begin{center}
      \includegraphics[scale=0.5]{files/c2.png}
    \end{center}

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement! So write it well because it could make a big difference.
In your statement, you have the opportunity to make  make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
%is important because if you need (GTA) funding, it will provide evidence
%that you can teach and communicate with students.

@@

@@

Having good grades is important, but unless your school is well-known, having top grades or rankings
usually will not help. This is simply because we cannot evaluate them.

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools (or don't want to go to the ones that you got it), try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., publish more papers, get more research experiences, etc.

You can also consider applying to MS programs, which are typically easier to get in (but you need to pay).  If you get in an MS at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is a common path for many students.

@@

%\subsection{R1, R2, ...}

\subsection{Selecting and Ranking Schools for CS}
   \begin{center}
    \includegraphics[scale=0.5]{files/c1.png}
   \end{center}

@@

\section{Choosing a Professor}

Now we discuss about finding advisers.

\subsection{Finding the right adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with, i.e., \emph{fits} you.  This is difficult and takes time.  Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\S\ref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\S\ref{sec:ta}), which gives you time to explore and find adviser.

@@

%It is worth noting that in many cases,  an application could be rejected by most reviewers, however have a "champion", who doesn't even have to be in the Admission comittee, that strongly advocate for your profile.

@@

\begin{center}
\includegraphics[scale=0.6]{files/c5.png}
\end{center}

If you're reaching out to a professor for the first time,  address them as either Prof. Lastname or Dr. Lastname. Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. It's not necessary, so stick with Prof. or Dr. Lastname.

@@

Note that in some universities the formal title \emph{Dr.} Lastname is preferred. You just need to observe and follow the conventions at your particular institution. Additionally, be aware that not all faculty members may hold a Ph.D., in which case using Prof. Lastname is a suitable alternative.

@@

%\section{Having fun during a PhD?}
%PhD students \emph{and faculty} probably find it amusing about the notion that students, especially international ones, can genuinely enjoy their PhD studies. In fact, after reading posts after posts on VietPhD.org on how PhD students are commonly mistreated, stressed, it seems being miserable is a norm during a PhD study.

%There are many advice on surviving PhD that you can follow. But here I just list a few that works for me and what I advice my students to do.\tvn{TODO}

\section{Will I be miserable during my a PhD?}\label{sec:happy}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable. Certainly issues including bad relationships with professors, conflicts with co-authors and lab mates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student) \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, happy students, etc.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

%\paragraph{Diversity} US universities prioritize diversity and inclusion. Students need to respect and appreciate varied opinions, backgrounds, and cultures. Unlike some countries where certain voices are marginalized, in the US, all perspectives are valued equally (especially at universities). Racism or discrimination will have serious consequences, including academic and disciplinary actions.

@@

Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on top CS conferences.
    \begin{table}
    \centering
    \small
    \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & MIT & 29 & Univ. of California - Riverside\\
    5 & Georgia Institute of Technology         & 30 & Northwestern University\\
    6 & Stanford University& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Univ. of California - Berkeley  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University\red{$^*$} &37&  Ohio State University\\
    13 & University of Wisconsin - Madison\red{$^*$}  &38& Texas A\&M University\red{$^*$} \\
    14 & Columbia University   &39 & Boston University  \\
    15 &   Purdue University  &40& North Carolina State University\\\  
    16 & University of Texas at Austin   &41 & University of Utah \\
    17 & University of Pennsylvania\red{$^*$} &42 & University at Buffalo\red{$^*$}\\
    18 & Princeton University  & 43& Rice University\\
    19 & University of Massachusetts-Amherst\red{$^*$} & 44&  University of Colorado-Boulder \\
    20 &  New York University  &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech\red{$^*$}  \\
    22 & University of Southern California &47&  Arizona State University\red{$^*$} \\
    23 & Stony Brook University\red{$^*$} &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina\red{$^*$} \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\paragraph{History} This document was conceived during a lunch with Craig Yu at GMU.  We talked on about why GMU were not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \S\ref{sec:ranking}). We wished there were a way for international students to know about the US PhD programs (as well as for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also  a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students in non-CS fields or not in US. Like ChatGPT, their answers to CS PhD program in the US could be informative and helpful, but unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

@@

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Prior to GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the Masters program in Software Engineering at GMU (thus also have some experience with the MS admission process--which is quite different than PhD). I got my undergraduate and MS degrees from Penn State. I completed my PhD at the University of New Mexico-Albuquerque and a postdoc at the University of Maryland-College Park.

>>>>

\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

@@

\makeatletter
\def\maketitle{%
  \null
  \thispagestyle{empty}%
  \vfill
  \begin{center}\leavevmode
    \normalfont
    {\LARGE\raggedright \textbf{\@title}\par}%
    \vfill%
    % \hrulefill\par
    {\Large \@author\par}%
    \vfill%
    {\large\raggedleft \@date\par}%
  \end{center}%
  \vfill
  \null
  \cleardoublepage
}
\makeatother
% \author{ThanhVu H. Nguyen}
% \title{TITLE OF BOOK}
% \date{}

@@

\chapter{Preface}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those in  smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines. 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and also likely universities in other countries).

@@

\chapter{Summary}\label{sec:summary}

Here, we outline the main points of the document. This gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

% \didi{Is there more information on typical strengths and common weaknesses of applications. This is especially useful to sophomore and junior students as they still have time to work on those strengths.}
% tvn: the main thing is research experiences

@@

% \begin{commentbox}[Hung]
%   The application fee varies between schools, and is about \$70--\$100. I found this number pretty ridiculous. In most cases, there is no proper justification for that amount. Yes, the application fee could act as a filter, but a nominal amount would do the job. While a blanket waiver might not be feasible, universities should extend the eligibility of fee waiver applications to more countries.
% \end{commentbox}

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and a lot better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \S\ref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \S\ref{sec:ranking}).

@@

Most CS PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, mentors) that know you well enough to talk about \emph{your specific research experience and capabilities}. This person can be a postdoc or even graduate student who has mentored you.

@@

\paragraph{Always Waive your Right to Review the LORs}  Choosing not to look at a reference letter is pretty standard in job and school applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writters, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

Reviewers might raise concerns (a red flag) about  a letter that is not waived, e.g., they will think that if you do not trust your letter writers, then you should find someone else to write for you. In short, it's a standard practice and a way of keeping things straightforward and respectful in the whole recommendation game.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the US. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters can be positive as they have helped many students getting into top PhD programs in the US.

@@

\begin{center}
  \includegraphics[scale=0.4]{files/c2.png}
\end{center}

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written LOR also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

@@

% \subsection{Outline of a LOR}\label{sec:lor-outline}

@@

Having good grades is important, but unless your school is well-known, having top grades or rankings
usually will not help because we cannot evaluate them.

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools (or don't want to go to the ones that admit you), try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc.

You can also consider applying to MS programs, which are typically easier to get in (but you need to pay).  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is a common path for many students.

@@

% \subsection{R1, R2, ...}

\subsection{Selecting and Ranking Schools for CS}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.5]{files/c1.png}
\end{center}

@@

\section{Choosing a Professor}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with, i.e., \emph{fits} you.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\S\ref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\S\ref{sec:ta}), which gives you time to explore and find adviser.

\subsection{Finding the right adviser}

@@

% It is worth noting that in many cases,  an application could be rejected by most reviewers, however have a "champion", who doesn't even have to be in the Admission comittee, that strongly advocate for your profile.

@@

\begin{center}
  \includegraphics[scale=0.5]{files/c5.png}
\end{center}

If you're reaching out to a professor for the first time,  address them as Prof. Lastname or Dr. Lastname. Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. It's not necessary, so stick with Prof. or Dr. Lastname.

@@

Note that in some universities the formal title Dr. Lastname is preferred over Prof. You just need to observe and follow the conventions at your particular institution. Additionally, be aware that not all faculty members may hold a Ph.D., in which case using Prof. Lastname is a suitable alternative.

@@

% \section{Having fun during a PhD?}
% PhD students \emph{and faculty} probably find it amusing about the notion that students, especially international ones, can genuinely enjoy their PhD studies. In fact, after reading posts after posts on VietPhD.org on how PhD students are commonly mistreated, stressed, it seems being miserable is a norm during a PhD study.

% There are many advice on surviving PhD that you can follow. But here I just list a few that works for me and what I advice my students to do.\tvn{TODO}

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories on how students are mistreated, stressed, and miserable. Certainly issues including bad relationships with professors, conflicts with co-authors and lab mates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student) \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, happy students, etc.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

% \paragraph{Diversity} US universities prioritize diversity and inclusion. Students need to respect and appreciate varied opinions, backgrounds, and cultures. Unlike some countries where certain voices are marginalized, in the US, all perspectives are valued equally (especially at universities). Racism or discrimination will have serious consequences, including academic and disciplinary actions.

@@

\begin{table}
  \centering
  \small
  \caption{Top 50 CS PhD programs in the U.S. (CSRankings, June 2023). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Univ. of California - Irvine \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 &  Duke University \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & MIT & 29 & Univ. of California - Riverside\\
    5 & Georgia Institute of Technology         & 30 & Northwestern University\\
    6 & Stanford University& 31 & Pennsylvania State University  \\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\  
    8 & University of Washington      &33 &  Harvard University \\
    9 &  Univ. of California - Berkeley  &34&  Univ. of California - Santa Cruz \\
    10 & Cornell University  & 35 &  Yale University \\
    11 & University of Maryland - College Park &  36& Brown University \\ 
    12 & Northeastern University\red{$^*$} &37&  Ohio State University\\
    13 & University of Wisconsin - Madison\red{$^*$}  &38& Texas A\&M University\red{$^*$} \\
    14 & Columbia University   &39 & Boston University  \\
    15 &   Purdue University  &40& North Carolina State University\\\  
    16 & University of Texas at Austin   &41 & University of Utah \\
    17 & University of Pennsylvania\red{$^*$} &42 & University at Buffalo\red{$^*$}\\
    18 & Princeton University  & 43& Rice University\\
    19 & University of Massachusetts-Amherst\red{$^*$} & 44&  University of Colorado-Boulder \\
    20 &  New York University  &45& University of Illinois at Chicago  \\
    21 & Univ. of California - Los Angeles &46& Virginia Tech\red{$^*$}  \\
    22 & University of Southern California &47&  Arizona State University\red{$^*$} \\
    23 & Stony Brook University\red{$^*$} &48&University of Minnesota \\
    24 & University of Chicago &49& University of Virginia \\
    25 & Univ. of California - Santa Barbara &50& University of North Carolina\red{$^*$} \\
    \bottomrule
  \end{tabular}
\end{table}

Tab.~\ref{tab:ranking} lists the top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on top CS conferences.

@@

\paragraph{History} This document was conceived during a lunch with Craig Yu at GMU.  We talked on about why GMU were not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \S\ref{sec:selecting-ranking-schools}). We wished there were a way for international students to know about the US PhD programs (as well as for US faculty to understand more about international students and therefore have better chance of recruiting and working with them). I was also  a member of the large VietPhD group on Facebook and saw many questions from students about PhD programs.  However, most active participant are students in non-CS fields or not in US. Like ChatGPT, their answers to questions on CS PhD program in the US could be informative and helpful, but unfortunately not always accurate and sometimes leading to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

@@

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Prior to GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have some experience with the MS admission process--which is quite different than PhD). %I got my undergraduate and MS degrees from Penn State. I completed my PhD at the University of New Mexico-Albuquerque and a postdoc at the University of Maryland-College Park.
